A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-09 latex/main.tex
Updates from Overleaf

\newcommand{\red}[1]{\textcolor{red}{#1}}

@@

\author[1,2,8,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
\author[1,7,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
\author[3,7,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
\author[4]{Katherine~Eaton}
\author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
\author[1,3,5,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
\author[4,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
\author[6,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

% NOTE: Trying to clarify that this of the pandemic period and
% we're not trying to cover *all* history.
The ARG is a comprehensive and deeply validated record of
the pandemic phase of SARS-CoV-2 evolutionary history,
accurately reflecting a range of results from the literature.
First, we compared the sc2ts ARG
with the state-of-the-art UShER tree inferred from the same dataset
by pruning both to their intersection of
2,475,418 samples and 27,507 sites (STAR Methods).
Eliminating recombination nodes and their descendants (which we expected to differ), 
we found very similar relationships among lineages in
the ARG and UShER tree based on tanglegrams
for a set of representative samples from 814 well-sampled lineages
(STAR Methods; Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
Overall, we also found close similarity between the ARG
and UShER tree in terms of parsimony 
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We compared the mutational spectra of major Variants of Concern (VOCs) in the ARG
against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We also evaluated how well the ARG reflects the structure of
the Pango nomenclature system
by computing Pango assignments for all 2.75 million nodes in the ARG
using Pangolin~\cite{OToole2021assignment} (STAR Methods).
Only a small minority of the exported sample alignments (21,680 ; 0.87\%) disagreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of the Pango lineages (1473 of 2058)
% SO: Not "these" as that seemed to refer to the conflicting nodes.
have monophyletic origins in the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
The ARG is time-resolved, with internal node dates estimated
by a non-parametric method~\cite{Pope2025} (STAR Methods).
We show that the dates agree well with
Nextstrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
and overall accuracy is comparable to 
Chronumental~\cite{sanderson2021chronumental} applied to 
an UShER tree (Figure~\ref{fig:node_time_testing},
Document~\ref{sec:evaluating_node_dates}).
Finally, we verified
that the origins of the
Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
and
Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
VOCs reflect expectations from the literature~\cite{RambautAlpha2020,ECDCreport,
Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.

@@

Further contributions of this work arise from 
the detailed study of the sc2ts ARG,
providing insights into the evolution of SARS-CoV-2
and the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X lineages, finding examples of
clustering of named lineages under a single shared recombination event.
Second, we used the ARG to study breakpoint intervals 
for recombination events, showing that the relationship between the 
size of the interval in base pairs and the divergence time of the parental lineages matches theoretical expectations.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
These results underscore the utility of the ARG, and 
the efficient and reproducible pipelines throughout this work 
provide a template for future analyses in which recombination
is accounted for by default.

@@

Alignments read from the VCF Zarr file were first preprocessed
to mask any non-nucleotide characters (A, C, G, or T) as missing data.
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
Then, excluding the 100 sites identified as problematic, we filtered any samples that had missing data at more than 500 sites (this was set to 10,000 sites for the initial period up to 2020-03-01
when sampling density was low).
% SO: Moved clause about 100 sites to the beginning so that it was clearer that 10,000 is in comparison to 500 not 100.
\subsubsection*{Li and Stephens model}

>>>>

@@

\author[1,2,8,*,\orcidlink{0000-0002-6351-8873}]{Shing~H.~Zhan}
\author[1,7,\orcidlink{0000-0002-3536-6411}]{Yan~Wong}
\author[3,7,\orcidlink{0000-0001-6402-9396}]{Anastasia~Ignatieva}
\author[4,\orcidlink{0000-0001-6862-7756}]{Katherine~Eaton}
\author[1,\orcidlink{0000-0002-3547-7226}]{Isobel~Guthrie}
\author[1,\orcidlink{0000-0002-1982-6801}]{Benjamin~Jeffery}
\author[1,3,5,\orcidlink{0000-0002-0824-0047}]{Duncan~S.~Palmer}
\author[4,\orcidlink{0000-0002-1543-4501}]{Carmen~Lia~Murall}
\author[6,\orcidlink{0000-0003-3042-0818}]{Sarah~P.~Otto}
\author[1,**,\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

% NOTE: Trying to clarify that this of the pandemic period and
% we're not trying to cover *all* history.
The ARG is a comprehensive and deeply validated record of
the pandemic phase of SARS-CoV-2 evolutionary history,
accurately reflecting a range of results from the literature.
First, we compared the sc2ts ARG
with the state-of-the-art UShER tree inferred from the same dataset
by pruning both to their intersection of
2,475,418 samples and 27,507 sites (STAR Methods).
Eliminating recombination nodes and their descendants (which we expected to differ), 
we found very similar relationships among lineages in
the ARG and UShER tree based on tanglegrams
for a set of representative samples from 814 well-sampled lineages
(STAR Methods; Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
Overall, we also found close similarity between the ARG
and UShER tree in terms of parsimony 
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We compared the mutational spectra of major Variants of Concern (VOCs) in the ARG
against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We also evaluated how well the ARG reflects the structure of
the Pango nomenclature system
by computing Pango assignments for all 2.75 million nodes in the ARG
using Pangolin~\cite{OToole2021assignment} (STAR Methods).
Only a small minority of the exported sample alignments (21,680 ; 0.87\%) disagreed
% NOTE Sally pointed out as potentially needing clarification.
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of the Pango lineages (1473 of 2058)
% SO: Not "these" as that seemed to refer to the conflicting nodes.
have monophyletic origins in the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
The ARG is time-resolved, with internal node dates estimated
by a non-parametric method~\cite{Pope2025} (STAR Methods).
We show that the dates agree well with
Nextstrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
and overall accuracy is comparable to 
Chronumental~\cite{sanderson2021chronumental} applied to 
an UShER tree (Figure~\ref{fig:node_time_testing},
Document~\ref{sec:evaluating_node_dates}).
Finally, we verified
that the origins of the
Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
and
Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
VOCs reflect expectations from the literature~\cite{RambautAlpha2020,ECDCreport,
Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.

@@

Further contributions of this work arise from 
the detailed study of the sc2ts ARG,
providing insights into the evolution of SARS-CoV-2
and the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X lineages, finding examples of
clustering of named lineages under a single shared recombination event.
Second, we used the ARG to study breakpoint intervals 
for recombination events, showing that the relationship between the 
size of the interval in base pairs and the divergence time of the parental lineages matches theoretical expectations.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
These results underscore the utility of the ARG 
and provide a template for future analyses.

@@

Alignments read from the VCF Zarr file were first preprocessed
to mask any non-nucleotide characters (A, C, G, or T) as missing data.
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
Then, excluding the 100 sites identified as problematic, we filtered any samples that had missing data at more than 500 sites (this was set to 10,000 sites for the initial period up to 2020-03-01
when sampling density was low).
% SO: Moved clause about 100 sites to the beginning so that it was clearer that 10,000 is in comparison to 500 not 100.

\subsubsection*{Li and Stephens model}
